---
title: "Project 2"
author: "Morgan Henderson, Preena Maruthavelu"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("knitr", "tidyr", "dplyr", "ggplot2", "ggthemes", "stringr")
install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(stringr)
```


# Group Project 2
## Repo Link:
https://github.com/hey-morgs/Group-Project-2.git

## Background
The Global Health Initiative has recently employed a new data science response team, CSIT-165, that uses data science to characterize pandemic diseases. CSIT-165 disseminates data driven analyses to global and local decision makers.

CSIT-165 is a conglomerate comprised of two fabricated entities: World Health Organization (WHO) and U.S. Pandemic Response Team (USPRT). 

Morgan is part of the WHO
Preena is part of the USPRT

### Objective 1
```{r ob1 - Morgan}

#load data
global_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
global_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
us_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
us_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
global_cases_df <- read.csv(global_cases, header = TRUE, na.strings = c("", " "))
global_deaths_df <- read.csv(global_deaths, header = TRUE, na.strings = c("", " "))
us_cases_df <- read.csv(us_cases, header = TRUE, na.strings = c("", " "))
us_deaths_df <- read.csv(us_deaths, header = TRUE, na.strings = c("", " "))

library("leaflet")
library("dplyr")

# Select the relevant columns for country/province, latitude, longitude, and the last date column
cases_cols <- c("Province.State", "Country.Region", "Lat", "Long", "X3.9.23")
deaths_cols <- c("Province.State", "Country.Region", "Lat", "Long", "X3.9.23")

# Prepare the confirmed cases data frame
cases_data <- global_cases_df %>%
  select(cases_cols) %>%
  group_by(Country.Region) %>%
  summarise(ConfirmedCases = sum(X3.9.23, na.rm = TRUE),
            MeanLat = mean(Lat, na.rm = TRUE),
            MeanLong = mean(Long, na.rm = TRUE))

# Prepare the deaths data frame
deaths_data <- global_deaths_df %>%
  select(deaths_cols) %>%
  group_by(Country.Region) %>%
  summarise(Deaths = sum(X3.9.23, na.rm = TRUE),
            MeanLat = mean(Lat, na.rm = TRUE),
            MeanLong = mean(Long, na.rm = TRUE))
str(cases_data)
str(deaths_data)
#print (deaths_data)
# Define color palette for cases
casesColor <- colorFactor(
  palette = c("blue", "purple", "red"),
  domain = quantile(cases_data$ConfirmedCases, probs = c(0.33, 0.67), na.rm = TRUE)
)
# Define color for deaths
deathsColor <- colorFactor(
  palette = c("blue", "purple", "red"),
  domain = quantile(deaths_data$Deaths, probs = c(0.33, 0.67), na.rm = TRUE)
)

map <- leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = cases_data, group = "Confirmed Cases",
                   lat = ~ MeanLat,
                   lng = ~ MeanLong,
                   color = ~casesColor(ConfirmedCases), 
                   label = ~sprintf('%s', Country.Region),
                   popup = ~sprintf('Cases: %s', ConfirmedCases),
                   radius = 6,
                   stroke = FALSE
                   ) %>%
  addCircleMarkers(data = deaths_data, group = "Deaths",
                   lat = ~ MeanLat,
                   lng = ~ MeanLong,
                   color = ~deathsColor(Deaths),
                   label = ~sprintf('%s', Country.Region),
                   popup = ~sprintf('Deaths: %s', Deaths),
                   radius = 6,
                   stroke = FALSE)%>%
addLayersControl(
  overlayGroups = c("Confirmed Cases", "Deaths"),
  options = layersControlOptions(collapsed = FALSE)
)

map
```

=======

```{r ob2 - Preena}
top_cases <- global_cases_df %>%
  group_by(Country.Region) %>% 
  summarize(Cases=sum(X3.9.23)) %>% 
  arrange(desc(Cases))

top_cases <- top_cases %>% mutate(Rank = 1:nrow(top_cases)) %>% select(Rank, everything())

top_deaths <- global_deaths_df %>%
  group_by(Country.Region) %>% 
  summarize(Deaths=sum(X3.9.23)) %>% 
  arrange(desc(Deaths))

top_deaths <- top_deaths %>% mutate(Rank = 1:nrow(top_deaths)) %>% select(Rank, everything())

consolidated <- full_join(top_cases, top_deaths, by ='Rank')

kable(consolidated)
  
```

