---
title: "Project 2"
author: "Morgan Henderson, Preena Maruthavelu"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("knitr", "tidyr", "dplyr", "ggplot2", "ggthemes", "stringr", "kableExtra")
install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(stringr)
library(kableExtra)
```


# Group Project 2
## Repo Link:
https://github.com/hey-morgs/Group-Project-2.git

## Background
The Global Health Initiative has recently employed a new data science response team, CSIT-165, that uses data science to characterize pandemic diseases. CSIT-165 disseminates data driven analyses to global and local decision makers.

CSIT-165 is a conglomerate comprised of two fabricated entities: World Health Organization (WHO) and U.S. Pandemic Response Team (USPRT). 

Morgan is part of the WHO
Preena is part of the USPRT

### Objective 1
```{r ob1 - Morgan}

#load data
global_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
global_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
us_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
us_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
global_cases_df <- read.csv(global_cases, header = TRUE, na.strings = c("", " "))
global_deaths_df <- read.csv(global_deaths, header = TRUE, na.strings = c("", " "))
us_cases_df <- read.csv(us_cases, header = TRUE, na.strings = c("", " "))
us_deaths_df <- read.csv(us_deaths, header = TRUE, na.strings = c("", " "))


```

```{r ob2 - Preena}
top_cases <- global_cases_df %>%
  group_by(Country.Region) %>% 
  summarize(Cases=sum(X3.9.23)) %>% 
  arrange(desc(Cases))

top_cases <- top_cases %>% mutate(Rank = 1:nrow(top_cases)) %>% select(Rank, everything())

top_deaths <- global_deaths_df %>%
  group_by(Country.Region) %>% 
  summarize(Deaths=sum(X3.9.23)) %>% 
  arrange(desc(Deaths))

top_deaths <- top_deaths %>% mutate(Rank = 1:nrow(top_deaths)) %>% select(Rank, everything())

consolidated <- full_join(top_cases, top_deaths, by ='Rank')

consolidated <- consolidated %>% rename("Top Cases Countries" = "Country.Region.x")
consolidated <- consolidated %>% rename("Top Deaths Countries" = "Country.Region.y")

kable(consolidated) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 1, "Confirmations" = 2, "Deaths" = 2))
  
```

