---
title: "Project 2"
author: "Morgan Henderson, Preena Maruthavelu"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("knitr", "tidyr", "dplyr", "ggplot2", "ggthemes", "stringr", "kableExtra")
install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(stringr)
library(RColorBrewer)
library(leaflet)
library(kableExtra)
library(cowplot)
```


# Group Project 2
## Repo Link:
https://github.com/hey-morgs/Group-Project-2.git

## Background
The Global Health Initiative has recently employed a new data science response team, CSIT-165, that uses data science to characterize pandemic diseases. CSIT-165 disseminates data driven analyses to global and local decision makers.

CSIT-165 is a conglomerate comprised of two fabricated entities: World Health Organization (WHO) and U.S. Pandemic Response Team (USPRT). 

Morgan is part of the WHO
Preena is part of the USPRT

### Objective 1

```{r ob1 - Morgan}

#load data
global_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
global_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
us_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
us_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
global_cases_df <- read.csv(global_cases, header = TRUE, na.strings = c("", " "))
global_deaths_df <- read.csv(global_deaths, header = TRUE, na.strings = c("", " "))
us_cases_df <- read.csv(us_cases, header = TRUE, na.strings = c("", " "))
us_deaths_df <- read.csv(us_deaths, header = TRUE, na.strings = c("", " "))

# Select the relevant columns for country/province, latitude, longitude, and the last date column
cases_cols <- c("Province.State", "Country.Region", "Lat", "Long", "X3.9.23")
deaths_cols <- c("Province.State", "Country.Region", "Lat", "Long", "X3.9.23")

# Prepare the confirmed cases data frame
cases_data <- global_cases_df %>%
  select(all_of(cases_cols)) %>%
  group_by(Country.Region) %>%
  summarise(ConfirmedCases = sum(X3.9.23, na.rm = TRUE),
            MeanLat = mean(Lat, na.rm = TRUE),
            MeanLong = mean(Long, na.rm = TRUE))

# Prepare the deaths data frame
deaths_data <- global_deaths_df %>%
  select(all_of(deaths_cols)) %>%
  group_by(Country.Region) %>%
  summarise(Deaths = sum(X3.9.23, na.rm = TRUE),
            MeanLat = mean(Lat, na.rm = TRUE),
            MeanLong = mean(Long, na.rm = TRUE))

# Define color palette for cases
casesColor <- colorNumeric(
  palette = "YlOrRd",
  domain = cases_data$ConfirmedCases
)
# Define color for deaths
deathsColor <- colorNumeric(
  palette = "YlOrRd",
  domain = deaths_data$Deaths
)
#create map
map <- leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = cases_data, group = "Confirmed Cases",
                   lat = ~ MeanLat,
                   lng = ~ MeanLong,
                   color = ~casesColor(ConfirmedCases), 
                   label = ~sprintf('%s', Country.Region),
                   popup = ~sprintf('Cases: %s', ConfirmedCases),
                   radius = 6,
                   stroke = FALSE,
                   fillOpacity = 0.8
                   ) %>%
  addCircleMarkers(data = deaths_data, group = "Deaths",
                   lat = ~ MeanLat,
                   lng = ~ MeanLong,
                   color = ~deathsColor(Deaths),
                   label = ~sprintf('%s', Country.Region),
                   popup = ~sprintf('Deaths: %s', Deaths),
                   radius = 6,
                   stroke = FALSE,
                   fillOpacity = 0.8)%>%
addLayersControl(
  overlayGroups = c("Confirmed Cases", "Deaths"),
  options = layersControlOptions(collapsed = FALSE)
)
#print map
map
```

### Objective 2

```{r ob2 - Preena}
top_cases <- global_cases_df %>%
  group_by(Country.Region) %>% 
  summarize(Cases=sum(X3.9.23)) %>% 
  arrange(desc(Cases))

top_cases <- top_cases %>% mutate(Rank = 1:nrow(top_cases)) %>% select(Rank, everything())

top_deaths <- global_deaths_df %>%
  group_by(Country.Region) %>% 
  summarize(Deaths=sum(X3.9.23)) %>% 
  arrange(desc(Deaths))

top_deaths <- top_deaths %>% mutate(Rank = 1:nrow(top_deaths)) %>% select(Rank, everything())

consolidated <- full_join(top_cases, top_deaths, by ='Rank')

consolidated <- consolidated %>% rename("Top Cases Countries" = "Country.Region.x")
consolidated <- consolidated %>% rename("Top Deaths Countries" = "Country.Region.y")

kable(consolidated) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 1, "Confirmations" = 2, "Deaths" = 2))
  
```

### Objective 3
```{r ob3 - Morgan}
# Filter California cases and remove unwanted columns
df_ca <- us_cases_df %>%
  filter(Province_State == "California") %>%
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Admin2, -Country_Region, -Lat, -Long_, -Combined_Key)

# Transform the data from wide to long format
df_ca_long <- df_ca %>%
  pivot_longer(cols = starts_with("X"), names_to = "Date", values_to = "Cases")

# Convert the date to Date class and sum the cases
df_ca_sum <- df_ca_long %>%
  mutate(Date = as.Date(gsub("X", "", Date), format = "%m.%d.%y")) %>%
  group_by(Date) %>%
  summarise(Cases = sum(Cases, na.rm = TRUE))

# Define dates of interest
start_date <- as.Date("2020-03-19")
vaccine_date <- as.Date("2021-01-29")
delta_date <- as.Date("2021-07-04")
omicron_date <- as.Date("2021-12-20")

# Plot the data
state_plot <- ggplot(df_ca_sum, aes(x = Date, y = Cases)) +
  geom_line(linewidth = 2) +
  geom_vline(xintercept = start_date, linetype = "dashed", color = "blue", linewidth = 1) +
  geom_vline(xintercept = vaccine_date, linetype = "dashed", color = "blue", linewidth = 1) +
  geom_vline(xintercept = delta_date, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = omicron_date, linetype = "dashed", color = "red", linewidth = 1) +
  geom_text(aes(x = start_date, y = 10000000, label = "SAH"), vjust = -0.5) +
  geom_text(aes(x = vaccine_date, y = 10000000, label = "EUA"), vjust = -0.5) +
  geom_text(aes(x = delta_date, y = 10000000, label = "Delta"), vjust = -0.5) +
  geom_text(aes(x = omicron_date, y = 10000000, label = "Omicron"), vjust = -0.5) +
  labs(x = "", y = "Confirmed Cases",
       title = "COVID-19 Confirmed Cases in California"
       ) +
  scale_x_date(
    breaks = c(start_date, vaccine_date, delta_date, omicron_date),
    labels = c("2020-03-19", "2021-01-29", "2021-07-04", "2021-12-20")
  )

# Filter California cases and remove extra columns
df_ca <- us_cases_df %>%
  filter(Province_State == "California") %>%
  select(Admin2, starts_with("X"))

# Transform the data from wide to long format
df_ca_long <- df_ca %>%
  pivot_longer(cols = starts_with("X"), names_to = "Date", values_to = "Cases")

# Convert the date to Date class and sum the cases
df_ca_sum <- df_ca_long %>%
  mutate(Date = as.Date(gsub("X", "", Date), format = "%m.%d.%y")) %>%
  group_by(Admin2, Date) %>%
  summarise(Cases = sum(Cases, na.rm = TRUE))

# Find the three cities with highest total cases
top_cities <- df_ca_sum %>%
  group_by(Admin2) %>%
  summarise(TotalCases = sum(Cases, na.rm = TRUE)) %>%
  top_n(3, TotalCases) %>%
  pull(Admin2)

# Filter the data for these cities
df_top_cities <- df_ca_sum %>% filter(Admin2 %in% top_cities)

label_data <- df_top_cities %>%
  group_by(Admin2) %>%
  filter(Date == as.Date("2022-10-01"))

# Define dates of interest
start_date <- as.Date("2020-03-19")
vaccine_date <- as.Date("2021-01-29")
delta_date <- as.Date("2021-07-04")
omicron_date <- as.Date("2021-12-20")

# Plot the data
city_plot <- ggplot(df_top_cities, aes(x = Date, y = Cases, color = Admin2)) +
  geom_line(linewidth = 2) +
  geom_text(data = label_data,
            aes(label = Admin2),
            color = "black",
            size = 3, 
            check_overlap = TRUE) +
  geom_vline(xintercept = start_date, linetype = "dashed", color = "blue", linewidth = 1) +
  geom_vline(xintercept = vaccine_date, linetype = "dashed", color = "blue", linewidth = 1) +
  geom_vline(xintercept = delta_date, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = omicron_date, linetype = "dashed", color = "red", linewidth = 1) +
  scale_color_manual(values = c("green", "purple", "orange"))+
  labs(y = "Cases",
       title = "",
       color = "City") +
  theme(legend.position = "none")
  scale_x_date(date_breaks = "1 year", 
               date_labels = "")

#combine plots using cowplot
combined_plot <- plot_grid(state_plot, city_plot, ncol = 1, align = "v")
plot_grid(combined_plot, ncol = 1, align = "v")

```

